A family of accessibility measures derived from spatial interaction principles

### 2025-04-06 paper/manuscript.qmd
code

unc_acc <- OD |>
  group_by(oid) |>
  summarize(V_unc_i_1 = sum(D_j * f1),
            V_unc_i_2 = sum(D_j * f2),
            V_unc_i_3 = sum(D_j * f3))

@@

However, we can compare values within the same travel behavior scenario across different zones but values are not inherently meaningful due to the units. For example, considering the most decaying scenario at zone 1, it captures `r unc_acc$V_unc_i_1[1] - unc_acc$V_unc_i_1[2]` more $f_1(c_{ij})$-weighted-physicians than zone 2, and `r unc_acc$V_unc_i_1[3] - unc_acc$V_unc_i_1[1]` fewer than zone 3. In the least decaying scenario, zone 1 captures `r unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_3[2]` more $f_3(c_{ij})$-weighted-physicians physicians than zone 2, and `r unc_acc$V_unc_i_3[3] - unc_acc$V_unc_i_3[1]` fewer than zone 3. Both scenarios at zone 1 captures an intermediate $V_i^0$ value between the two other zones, but at different ratios: the least decaying scenario captures `r unc_acc$V_unc_i_1[1]/unc_acc$V_unc_i_1[2]` and `r unc_acc$V_unc_i_1[1]/unc_acc$V_unc_i_1[3]` times the $f_1(c_{ij})$-weighted-physicians than zone 2 and 3 respectively, while the most decaying scenario captures `r unc_acc$V_unc_i_3[1]/unc_acc$V_unc_i_3[2]` and `r unc_acc$V_unc_i_3[1]/unc_acc$V_unc_i_3[3]` times the $f_3(c_{ij})$-weighted-physicians. The ratios between the zones in the two scenarios are different. These differences are tied up in the travel impedance functions, and hence the unconstrained accessibility values themselves should not be compared. <!--maybe a better way of saying this?--> This simple example illustrates that the differences and ratios of unconstrained accessibility values are not inherently meaningful. While one could attempt to adjust the units post-calculation (e.g., scaling, population normalization) or select comparable impedance functions (at the cost of accurately reflecting travel behavior), such adjustments may introduce bias.

@@

A total constraint proportionally adjusts zonal accessibility values based on the total population and/or opportunities in the region. The introduction of this constraint introduces a form of competition, as all zonal values are a proportion of the system total, be it the regional opportunities or regional population depending on the case. We define two cases: a *totally-constrained opportunity* accessibility case, interpreted as Hansen's accessibility with a constraint, and a *totally-constrained population* accessibility case, interpreted as constrained market potential.

@@

@@

\noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination ($V^T_i = \kappa^T\sum_j W^{(2)}_j$). The role of $\kappa^T$ in this reformulation of accessibility is to adjust the number of opportunities accessible from $i$ so that they represent a proportion of the total number of opportunities in the region. Constant $\kappa^T$ then assigns opportunities in proportion to the impedance between $i$ and $j$. For this reason, we refer to it as a proportional allocation factor.

@@

>>>>

unc_acc_ij <- OD |>
  group_by(oid) |>
  reframe(V_unc_ij_1 = D_j * f1,
          V_unc_ij_2 = D_j * f2,
          V_unc_ij_3 = D_j * f3)

unc_acc <- unc_acc_ij |>
  group_by(oid) |>
  summarize(V_unc_i_1 = sum(V_unc_ij_1),
            V_unc_i_2 = sum(V_unc_ij_2),
            V_unc_i_3 = sum(V_unc_ij_3))

## unconstrained access in one step
# unc_acc <- OD |>
#   group_by(oid) |>
#   summarize(V_unc_i_1 = sum(D_j * f1),
#             V_unc_i_2 = sum(D_j * f2),
#             V_unc_i_3 = sum(D_j * f3))

@@

However, we can compare values within the same travel behavior scenario across different zones but values are not inherently meaningful due to the units. For example, considering the most decaying scenario at zone 1, it captures `r unc_acc$V_unc_i_1[1] - unc_acc$V_unc_i_1[2]` more $f_1(c_{ij})$-weighted-physicians than zone 2, and `r unc_acc$V_unc_i_1[3] - unc_acc$V_unc_i_1[1]` fewer than zone 3. In the least decaying scenario, zone 1 captures `r unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_3[2]` more $f_3(c_{ij})$-weighted-physicians physicians than zone 2, and `r unc_acc$V_unc_i_3[3] - unc_acc$V_unc_i_3[1]` fewer than zone 3. Both scenarios at zone 1 captures an intermediate accessibility value $V_i^0$ between the two other zones, but at different ratios: the least decaying scenario captures `r unc_acc$V_unc_i_1[1]/unc_acc$V_unc_i_1[2]` and `r unc_acc$V_unc_i_1[1]/unc_acc$V_unc_i_1[3]` times the $f_1(c_{ij})$-weighted-physicians than zone 2 and 3 respectively, while the most decaying scenario captures `r unc_acc$V_unc_i_3[1]/unc_acc$V_unc_i_3[2]` and `r unc_acc$V_unc_i_3[1]/unc_acc$V_unc_i_3[3]` times the $f_3(c_{ij})$-weighted-physicians. The ratios between the zones in the two scenarios are different. These differences are tied up in the travel impedance functions, and hence the unconstrained accessibility values themselves should not be compared. <!--maybe a better way of saying this?--> This simple example illustrates that the differences and ratios of unconstrained accessibility values are not inherently meaningful. While one could attempt to adjust the units post-calculation (e.g., scaling, population normalization) or select comparable impedance functions (at the cost of accurately reflecting travel behavior), such adjustments may introduce bias.

@@

A total constraint proportionally adjusts zonal accessibility values based on the total population and/or opportunities in the region. The introduction of this constraint introduces a form of competition, as all zonal values are a proportion of the system total, be it the regional opportunities or regional population depending on the case. We define two cases: (a) a case where accessibility is constrained by the total number of opportunities (aka *totally-constrained opportunity*) and which is interpreted as Hansen's accessibility with a constraint, and (b) a case where accessibility is constrained by the total population (aka *totally-constrained population*) and which is interpreted as constrained market potential.

@@

<!-- rafa: do we need to keep equations 31 and 32 -->

@@

\noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination ($V^T_i = \kappa^T\sum_j W^{(2)}_j$). The role of $\kappa^T$ in this reformulation of accessibility is to transform between units and to adjust the number of opportunities accessible from $i$ so that they represent a proportion of the total number of opportunities in the region. Constant $\kappa^T$ then assigns opportunities in proportion to the impedance between $i$ and $j$. For this reason, we refer to it as a proportional allocation factor.

@@

<!-- rafa: 'The role of $\kappa^T$ in this reformulation of accessibility is to..'. This sentence is gold and it should be said earlier on. My personal preference is that first we need to explain to the reader in plain language what the constraint is, and what it means in practical terms (why it matters). Then we can show this with the equations.  -->
